Coherent Hyper-Raman Four Wave Mixing Vibrational Spectroscopy

### 2024-07-09 paper.tex
edits

\begin{figure*}[!htbp]
	\centering
	\includegraphics[width=6.66in]{f1.png}
	\caption{Wave Mixing Energy Level (WMEL) diagrams of difference frequency generation (DFG), spontaneous Raman scattering, hyper difference frequency generation (HDFG) and spontaneous hyper-Raman scattering. \cite{RN286, RN352}
		$\ket{g}$ is the ground state, $\ket{v}$ is an infrared active vibration and $\ket{m}, \ket{n}$ are virtual states.
		Solid and dashed horizontal lines indicate real and virtual states, whereas solid and dotted arrows indicate ket and bra side transitions, respectively. %todo: add Raman to this WMEL! dfg Raman hyperDFG hyperRaman 
		}
	\label{fig:comparisonwmel}
\end{figure*}

@@

To make the connection between HDFG and hyper-Raman scattering, we investigate its gross selection rules.
By propagating density matrix elements in the steady state limit, the HDFG hyperpolarizability is \cite{RN119}
\begin{equation}\label{sivegamma}
		\gamma_{ijkl} =	- \sum_{m, n} \frac{1}{\varepsilon_0} \frac{1}{4D} \frac{1}{\hbar^3} \frac{\mu^{vn}_{i} \mu^{nm}_{j} \mu^{mg}_{k} \mu^{gv}_{l} }{\Delta_{nv} \Delta_{mv}\Delta_{gv}}  \rho_{gg}
\end{equation}
where: $\mu^{ab}_{j}$ is the $j^{th}$ element of $\mel{a}{\vec{\mu}}{b}$, $\Delta_{kl} = \omega_{kl} - \omega_{j} - i\Gamma_{kl}$, $\omega_j$ is the frequency of the j$^{th}$ input field, $\Gamma_{kl}$ is the dephasing of $\rho_{kl}$ and $\rho_{gg}$ is the ground state population.
D, the Maker-Terhune degeneracy factor, accounts for permutation symmetry in $\gamma_{ijkl}$.\cite{RN134} 
For a HDFG experiment using two or three distinct input fields, as considered here, D = 6.
Note that $\ket{m}, \ket{n}$ are virtual states (\autoref{fig:sivewmel}).
By contracting over the virtual states to form the hyper-Raman hyperpolarizability $\beta$, this expression can be simplified immensely.\cite{Long1970} 
Note, however, that if $\omega_2$, $\omega_3$, as depicted in \autoref{fig:comparisonwmel}, are nearly resonant with any state, the contraction is not allowed. \cite{Placzek1934, Long1970}
Assuming $\omega_2$, $\omega_3$ are significantly detuned from any local resonance, \autoref{sivegamma} is written as 
\begin{equation}\label{sivebeta}
	\gamma_{ijkl} =	-\frac{1}{\varepsilon_0} \frac{1}{24 \hbar}\frac{\beta^{gv}_{ijk} \mu^{gv}_{l}}{\Delta_{gv}} \rho_{gg}
\end{equation}
Taylor expanding the dipole and first hyperpolarizability operators in terms of an n$^{\text{th}}$ normal mode coordinate $Q_n$ about equilibrium: \cite{Long1970}
\begin{subequations}
	\begin{equation}
		\mu = \mu_0 + \frac{\partial \mu}{\partial Q_n} Q_n + \order{{Q_m Q_n}}
	\end{equation}
	\begin{equation}
		\beta = \beta_{0} + \frac{\partial \beta}{\partial Q_n} Q_n + \order{{Q_m Q_n}}
	\end{equation}
\end{subequations}
Substituting into \autoref{sivebeta} gives the HDFG hyperpolarizability to $\order{Q_n}$ as \begin{equation}\label{SIVEselection}
	\gamma_{ijkl} =	-\frac{1}{\varepsilon_0} \frac{1}{48 m_n \omega_{vg}}  \frac{1}{{\Delta_{gv}}} \ \frac{\partial \beta^{gv}_{ijk}}{\partial Q_n} {\frac{\partial \mu^{gv}_{l}}{\partial Q_n}}  \rho_{gg}
\end{equation}
where $\mel{v}{Q_n}{g} = \sqrt{\frac{\hbar}{2m_n\omega_{vg}}}$ under the harmonic approximation,\cite{RN230} and $m_n$ is the ``mass'' of the $Q_n$ mode. 
Since this expression is non-zero in the harmonic oscillator limit, HDFG output is allowed for harmonic transitions. 
Similar to SFG, HDFG does not demand anharmonic ground state potential energy surfaces for spectral output. \cite{Shen94, Cho2000}
This selection rule is generally valid for any HDFG or hyper sum frequency generation (HSFG) process in the $-\vec{k}_1 + \vec{k}_2  + \vec{k}_3$ and $\vec{k}_1 + \vec{k}_2  + \vec{k}_3$ phasematching geometries when $\omega_2$ and $\omega_3$ are sufficiently detuned from any resonance.

@@

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=3.375in]{wmel3.png}
	\caption{WMEL diagrams of some (I-a,I-b) DR-HDFG and (II-a,II-b) DR-HSFG pathways. $\ket{g,0}$ denotes the ground states, $\ket{g,v}$ denotes a vibrational state on the ground electronic manifold; $\ket{m,n}$ denotes a virtual state, and $\ket{e,0}, \ket{e,v'}$ denote vibronic states.}
	\label{fig:doubsive}
\end{figure}

@@

It is well known that nonlinear wave mixing processes which share the same time ordering and output frequency will interfere. \cite{RN135, Bonn2024}
Understanding interfering pathways is useful because they can, in some cases, eliminate output. \cite{RN342, RN135}
The WMEL diagrams presented in \autoref{fig:sivewmel} outline resonant HDFG processes when the infrared pulse $(\omega_1)$ is resonant with a vibrational mode. 
However, if $\omega_2$ ($\omega_1$) becomes resonant (non-resonant), then other HDFG pathways appear.\cite{McDonnell2024} 
While the methods in \autoref{fig:sivewmel2} possess identical selection rules to \autoref{fig:sivewmel}, the WMEL diagrams presented in \autoref{fig:sivewmel2} may significantly interfere with DOVE-IR and DOVE-Raman spectra.
The time ordering in DOVE-IR and DOVE-Raman spectra is used to interpret anharmonicities, as DOVE-IR and DOVE-Raman destructively interfere when outputs are at the same frequency. \cite{RN135, RN324}
It is therefore important to understand how the corresponding HDFG pathways interfere.
To this end, we investigate how the HDFG-IR-II and HDFG-Raman pathways (\autoref{fig:sivewmel2}a, b, respectively) interfere under the rotating wave approximation.

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=3.375in]{figures/wmel2.png}
	\caption{Wave Mixing Energy Level (WMEL) diagrams of interfering hyper-DFG pathways in the  $\vec{k}_4 = -\vec{k}_1 + \vec{k}_2 + \vec{k}_3$ phasematching geometry. 
		(a) is similar to the process diagrammed in \autoref{fig:sivewmel}, whereas (b) is similar to a singly resonant coherent anti-Stokes Raman-type process.
		%todo: add pump-SIVE-probe to this set of WMELs
		}
	\label{fig:sivewmel2}
\end{figure}

We employ the response function formalism of Mukamel for this analysis. \cite{RN287}
The third order response functions, $R^{(3)}$, for \autoref{fig:sivewmel2}a ($R^{(3)}_{a}$) and \autoref{fig:sivewmel2}b ($R^{(3)}_{b}$) are 
\begin{subequations}
	\begin{equation} \label{mixing:a}
		R^{(3)}_{a} (t_2) = -\frac{i}{\hbar} [\beta(t_2), \mu(0)] \exp(-i\Delta_{v''g}t_2) \theta(t_2)
	\end{equation}
	\begin{equation}\label{mixing:b}
		\begin{split}
			R^{(3)}_{b} (t_2) & = \frac{i}{\hbar} [\beta(t_2), \mu(0)] \exp(-i\Delta_{v''g}t_2) \theta(t_2)\\
		\end{split}
	\end{equation}
\end{subequations}
The total response function is $R^{(3)}_\text{tot} (t_2) = R^{(3)}_{a} (t_2) + R^{(3)}_{b} (t_2) = 0$. 
Since the total output response function vanishes in this time-ordering, the pathways in \autoref{fig:sivewmel2} perfectly destructively interfere.
Therefore, to measure coherent dynamics via hyper-DFG, the time ordering shown in \autoref{fig:sivewmel} should be used.
These ideas have been seen experimentally in ultrafast DOVE work, as described elsewhere. \cite{RN367, McDonnell2024}
In the case of the time ordered (\autoref{fig:sivewmel}a) pathway, there are no opportunities for interference as all other pathways are nonresonant with the first interaction. 
A possible mechanism for interfering output in the (\autoref{fig:sivewmel}a) pathway are cascaded second order processes (SFG, DFG). \cite{RN243, RN300}
In isotropic media, SFG and DFG vanish under the electric dipole approximation,\cite{RN231} so cascades are unlikely to inhibit output in centrosymmetric media.
In media where SFG and DFG are allowed, e.g. surfaces or noncentrosymmetric media, second order cascades may become important.

>>>>

\begin{figure*}[!htbp]
	\centering
	\includegraphics[width=6.66in]{comparisonwmel.png}
	\caption{Wave Mixing Energy Level (WMEL) diagrams of difference frequency generation (DFG), spontaneous Raman scattering, hyper difference frequency generation (HDFG) and spontaneous hyper-Raman scattering. \cite{RN286, RN352}
		$\ket{g}$ is the ground state, $\ket{v}$ is an infrared active vibration and $\ket{m}, \ket{n}$ are virtual states.
		Solid and dashed horizontal lines indicate real and virtual states, whereas solid and dotted arrows indicate ket and bra side transitions, respectively. 
		}
	\label{fig:comparisonwmel}
\end{figure*}

@@

To make the connection between HDFG and hyper-Raman scattering, we investigate its gross selection rules.
By propagating density matrix elements in the steady state limit, the HDFG hyperpolarizability is \cite{RN119}
\begin{equation}\label{sivegamma}
		\gamma_{ijkl} =	- \sum_{m, n} \frac{1}{\varepsilon_0} \frac{1}{4D} \frac{1}{\hbar^3} \frac{\mu^{vn}_{i} \mu^{nm}_{j} \mu^{mg}_{k} \mu^{gv}_{l} }{\Delta_{nv} \Delta_{mv}\Delta_{gv}}  \rho_{gg}
\end{equation}
where: $\mu^{ab}_{j}$ is the $j^{th}$ element of $\mel{a}{\vec{\mu}}{b}$, $\Delta_{kl} = \omega_{kl} - \omega_{j} - i\Gamma_{kl}$, $\omega_j$ is the frequency of the j$^{th}$ input field, $\Gamma_{kl}$ is the dephasing of $\rho_{kl}$ and $\rho_{gg}$ is the ground state population.
D, the Maker-Terhune degeneracy factor, accounts for permutation symmetry in $\gamma_{ijkl}$.\cite{RN134} 
For a HDFG experiment using two or three distinct input fields, as considered here, D = 6.
By contracting over the virtual states to form the hyper-Raman hyperpolarizability $\beta$, this expression can be simplified immensely.\cite{Long1970} 
Note, however, that if $\omega_2$, $\omega_3$, as depicted in \autoref{fig:comparisonwmel}, are nearly resonant with any state, the contraction is not allowed. \cite{Placzek1934, Long1970}
Assuming $\omega_2$, $\omega_3$ are significantly detuned from any local resonance, \autoref{sivegamma} is written as 
\begin{equation}\label{sivebeta}
	\gamma_{ijkl} =	-\frac{1}{\varepsilon_0} \frac{1}{24 \hbar}\frac{\beta^{gv}_{ijk} \mu^{gv}_{l}}{\Delta_{gv}} \rho_{gg}
\end{equation}
Taylor expanding the dipole and first hyperpolarizability operators in terms of an n$^{\text{th}}$ normal mode coordinate $Q_n$ about equilibrium: \cite{Long1970}
\begin{subequations}
	\begin{equation}
		\mu = \mu_0 + \frac{\partial \mu}{\partial Q_n} Q_n + \order{{Q_m Q_n}}
	\end{equation}
	\begin{equation}
		\beta = \beta_{0} + \frac{\partial \beta}{\partial Q_n} Q_n + \order{{Q_m Q_n}}
	\end{equation}
\end{subequations}
Substituting into \autoref{sivebeta} gives the HDFG hyperpolarizability to $\order{Q_n}$ as \begin{equation}\label{SIVEselection}
	\gamma_{ijkl} =	-\frac{1}{\varepsilon_0} \frac{1}{48 m_n \omega_{vg}}  \frac{1}{{\Delta_{gv}}} \ \frac{\partial \beta^{gv}_{ijk}}{\partial Q_n} {\frac{\partial \mu^{gv}_{l}}{\partial Q_n}}  \rho_{gg}
\end{equation}
where $\mel{v}{Q_n}{g} = \sqrt{\frac{\hbar}{2m_n\omega_{vg}}}$ under the harmonic approximation,\cite{RN230} and $m_n$ is the ``mass'' of the $Q_n$ mode. 
Since this expression is non-zero in the harmonic oscillator limit, HDFG output is allowed for harmonic transitions. 
Similar to SFG, HDFG does not demand anharmonic ground state potential energy surfaces for spectral output. \cite{Shen94, Cho2000}
This selection rule is generally valid for any HDFG or hyper sum frequency generation (HSFG) process in the $-\vec{k}_1 + \vec{k}_2  + \vec{k}_3$ and $\vec{k}_1 + \vec{k}_2  + \vec{k}_3$ phasematching geometries when $\omega_2$ and $\omega_3$ are sufficiently detuned from any resonance.

@@

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=3.375in]{drwmel.png}
	\caption{WMEL diagrams of some (I-a,I-b) DR-HDFG and (II-a,II-b) DR-HSFG pathways. $\ket{g,0}$ denotes the ground states, $\ket{g,v}$ denotes a vibrational state on the ground electronic manifold; $\ket{m,n}$ denotes a virtual state, and $\ket{e,0}, \ket{e,v'}$ denote vibronic states.}
	\label{fig:doubsive}
\end{figure}

@@

It is well known that nonlinear wave mixing processes which share the same time ordering and output frequency will interfere. \cite{RN135, Bonn2024}
Understanding interfering pathways is useful because they can, in some cases, eliminate output. \cite{RN342, RN135}
The WMEL diagram presented in \autoref{fig:comparisonwmel} outlines a resonant HDFG process where the infrared pulse $(\omega_1)$ is resonant with a vibrational mode. 
However, if $\omega_2$ ($\omega_1$) becomes resonant (non-resonant), then other HDFG pathways appear.\cite{McDonnell2024} 
While the methods in \autoref{fig:sivewmel2} possess identical selection rules to \autoref{fig:sivewmel}, the WMEL diagrams presented in \autoref{fig:sivewmel2} will interfere.
\begin{figure}[!htbp]
	\centering
	\includegraphics[width=3.375in]{figures/timeorderedwmel.png}
	\caption{Wave Mixing Energy Level (WMEL) diagrams of interfering hyper-DFG pathways in the  $\vec{k}_4 = -\vec{k}_1 + \vec{k}_2 + \vec{k}_3$ phasematching geometry. 
		(a) is similar to the process diagrammed in \autoref{fig:sivewmel}, whereas (b) is similar to a singly resonant coherent anti-Stokes Raman-type process.
		%todo: add pump-SIVE-probe to this set of WMELs
		}
	\label{fig:sivewmel2}
\end{figure}

To investigate interfering pathways, we employ the response function formalism of Mukamel. \cite{RN287}
The third order response functions for \autoref{fig:sivewmel2}a ($R^{(3)}_{a}$) and \autoref{fig:sivewmel2}b ($R^{(3)}_{b}$) are 
\begin{subequations}
	\begin{equation} \label{mixing:a}
		R^{(3)}_{a} (t_2) = -\frac{i}{\hbar} [\beta(t_2), \mu(0)] \exp(-i\Delta_{v''g}t_2) \theta(t_2)
	\end{equation}
	\begin{equation}\label{mixing:b}
		\begin{split}
			R^{(3)}_{b} (t_2) & = \frac{i}{\hbar} [\beta(t_2), \mu(0)] \exp(-i\Delta_{v''g}t_2) \theta(t_2)\\
		\end{split}
	\end{equation}
\end{subequations}
The total response function is therefore $R^{(3)}_\text{tot} (t_2) = R^{(3)}_{a} (t_2) + R^{(3)}_{b} (t_2) = 0$. 
Since the total output response function vanishes in this time-ordering, the pathways in \autoref{fig:sivewmel2} perfectly destructively interfere.
Therefore, to measure coherent dynamics via hyper-DFG, the time ordering shown in \autoref{fig:comparisonwmel} should be used.
These ideas have been seen experimentally in ultrafast DOVE work, as described elsewhere. \cite{RN367, McDonnell2024}
In the case of the \autoref{fig:comparisonwmel} pathway, there are no opportunities for interference, as all other pathways are nonresonant with the first interaction. 
A possible mechanism for interfering output in the \autoref{fig:comparisonwmel} pathway are cascaded second order processes (SFG, DFG). \cite{RN243, RN300}
In isotropic media, SFG and DFG vanish under the electric dipole approximation,\cite{RN231} so cascades are unlikely to inhibit output in centrosymmetric media.
In media where SFG and DFG are allowed, e.g. surfaces or noncentrosymmetric media, second order cascades may become important.
